Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-20 papers/contact/paper.tex
minor cleanups in contact paper Tested everything on node05.

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by: \fixme{Which of the following two equations is more clear
  to you? We only want one.}
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{\sigma n_2}{6}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{\sigma n_2}{6}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.

>>>>

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by:
% \fixme{Which of the following two equations is more clear to you? We
%  only want one.}
\begin{align}
  %% g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
  %%   + \frac32 \frac{\sigma n_2}{6}\frac{\zeta}{(1-n_3)^2}
  %%   + \frac12 \frac{\left(\frac{\sigma n_2}{6}\right)^2 \zeta}{(1-n_3)^3}
  %%   \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.
